The influence of Gaia-Sausage-Enceladus-like merger events on the chemo-dynamics of galaxy discs

### 2023-03-07 src/tex/ms.tex
[showyourwork] overleaf sync

\begin{figure}
    \script{gas_profile_all.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/gas_profile_2d.pdf}
        \caption{
            Relative change in cold gas surface density profile before and after the merger events. Outside of a radius of $\sim5-10$ kpc the merging satellites contribute significant amounts of cold gas up to 200\%.
        }
        \label{fig:surf_den}
    \end{centering}
\end{figure}

@@

A sudden steepening of the galactic metallicity gradient is unexpected and has been deemed not possible (reference??? Ivan, any clues?). The physical reason for a steepening of the metallicity gradient might be a dilution of the outer disk metallicity by the fresh, unenriched gas brought in by the merging satellite, the continued or preferential enrichment of the galactic center or a combination of both. A dilution scenario might be expected from the general lower metallicity of low mass satellites compared to MW-mass galaxies owing to the shape of the stellar mass-metallicity relation and has previously been suggested for the creation of the chemical bimodality of the MW's stellar disk \citep[e.g.][]{Buck2020}.

In order to investigate the cause of the strong steepening of the metallicity gradient in our simulations we first plot in Figure~\ref{fig:surf_den} where in the disk the accreted gas mostly contributes. Figure~\ref{fig:surf_den} plots the relative change of cold gas surface mass density radial profile before the gas rich merger and after the merger finished as indicated by the gray lines in Figure~\ref{fig:half_mass}. This figure reveals an interesting finding. Consistently across all four galaxies the surface mass density of cold gas in the inner $5$ kpc is roughly constant or decreases slightly. Outside of $5$ kpc the surface mass density increases by $\sim75$ \% up to $\sim200$\%.

@@

\begin{figure}
    \script{feh_gradient.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/feh_gradient.pdf}
        \caption{
            Metallicity gradient of the cold gas disk during the merger events. We find that the contribution of additional metal poor gas lowers the metallicity in the outskirts of the disk effectively steepening the metallicity gradient by $\sim 0.1$ dex.
        }
        \label{fig:feh_grad}
    \end{centering}
\end{figure}

>>>>

\begin{figure}
    \script{gas_profile_all.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/gas_profile_2d.pdf}
        \caption{
            Relative change in cold gas surface density profile before and after the merger events as indicated in Fig.~\ref{fig:merger_ratio} by the gray lines. Outside of a radius of $\sim5-10$ kpc the merging satellites contribute significant amounts of cold gas up to 200\%.
        }
        \label{fig:surf_den}
    \end{centering}
\end{figure}

@@

A sudden steepening of the galactic metallicity gradient is unexpected and has been deemed not possible (reference??? Ivan, any clues?). The physical reason for a steepening of the metallicity gradient might be a dilution of the outer disk metallicity by the fresh, un-enriched gas brought in by the merging satellite, the continued or preferential enrichment of the galactic center or a combination of both. A dilution scenario might be expected from the general lower metallicity of low mass satellites compared to MW-mass galaxies owing to the shape of the stellar mass-metallicity relation and has previously been suggested for the creation of the chemical bimodality of the MW's stellar disk \citep[e.g.][]{Buck2020}.

In order to investigate the cause of the strong steepening of the metallicity gradient in our simulations we first plot in Figure~\ref{fig:surf_den} where in the disk the accreted gas mostly contributes. Figure~\ref{fig:surf_den} plots the relative change of cold gas surface mass density radial profile before the gas rich merger and after the merger finished as indicated by the gray lines in Figure~\ref{fig:half_mass}. This figure reveals an interesting finding. Consistently across all four galaxies the surface mass density of cold gas in the inner $5$ kpc is roughly constant or decreases slightly due to gas consumption by star formation and gas expulsion by feedback. Outside of $5$ kpc the surface mass density increases by $\sim75$ \% up to $\sim200$\% due to cold gas added byt he merger event, either by direct accretion of cold gas from the merging satellite or by triggered gas accretion from the circum-galactic medium due to tidal forces exerted by the interacting satellite.

Three scenarios can now lead to a steepenig of the metallicity gradient: (i) the metallicity of the star forming gas in the outskirts of the galaxies is diluted by the freshly accreted gas; (ii) the central parts of the galaxy keep enriching more strongly than the outskirts or (iii) a mix of both.
In fact, as was shown in \citet{Buck2020} in their Fig. 8 scenario (iii) is happening in these galaxies. For most of the time the ISM/cold gas metallicity evolution is self-similar and almost monotonically increasing at each radius but with markedly dilution events that correlate with merger events. In order to investigate this in more depth and study its effect on the metallicity gradient, we plot in Fig.~\ref{fig:feh_evolution} the difference in metallicity of the cold gas, $\Delta\mathrm{[Fe/H]}=\mathrm{[Fe/H]}(R<2\, \mathrm{kpc})-\mathrm{[Fe/H]}(R)$, between the gas in a given annulus at a radius $R$ and the metallicity in the central parts ($R<2$ kpc) of the galaxy, $\mathrm{[Fe/H]}(R<2\, \mathrm{kpc})$ as a function of time. We normalize this difference in metallicity, $\Delta\mathrm{[Fe/H]}$, by the central metallicity in order to take out the continued enrichment in [Fe/H] and highlight deviations from the self-similar evolution which is indicating a steepening or flattening of the metallicity gradient as measured over different radial ranges. Thus, a flat line in Fig.~\ref{fig:feh_evolution} would indicate perfect self-similar evolution while a more negative value indicates that a certain radius lacks behind the central galaxy in [Fe/H] enrichment and thus a steepening of the metallicity gradient. On the other hand, a progressively more positive or less negative value shows that the radius at hand enriches more strongly in metallicity than the central parts an thus indicates a flattening of the metallicity gradient.

@@

%\begin{figure}
%    \script{feh_gradient.py}
%    \begin{centering}
%        \includegraphics[width=\linewidth]{figures/feh_gradient.pdf}
%        \caption{
%            Metallicity gradient of the cold gas disk during the merger events. We find that the contribution of additional metal poor gas lowers the metallicity in the outskirts of the disk effectively steepening the metallicity gradient by $\sim 0.1$ dex.
%        }
%        \label{fig:feh_grad}
%    \end{centering}
%\end{figure}
